---
title: "Stan code for models in the R package `shrinkage`"
author: "Gwenaël GR Leday"
date: "16-11-2022"
output: rmarkdown::html_vignette
header-includes:
- \renewcommand{\textfraction}{0}
bibliography: refs.bib
biblio-style: "apalike"
link-citations: true
vignette: >
  %\VignetteIndexEntry{Stan}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>")
```

<!----------------------------------------------------------------------------->
<!--------------------------------- GLOBAL SHRINKAGE -------------------------->
<!----------------------------------------------------------------------------->
## Regression models with global shrinkage priors

### Gaussian model

The following provides stan code for fitting linear regression models with
various global shrinkage priors.

```{r eval=FALSE}
# stan code for global shrinkage with inverse-Gamma prior
code_brg_invGamma <- "
  data { 
    int<lower=0> p;
    int<lower=0> n;
    matrix[n,p] X;
    real y[n];
    real<lower=0> a;
    real<lower=0> b;
  }
  parameters {
    vector[p] beta;
    real<lower=0> sigma2;
    real<lower=0> tau2;
  }
  model {
      tau2 ~ inv_gamma(a, b);
      for(i in 1:p){
        beta[i] ~ normal(0, sqrt(sigma2*tau2));
      }
      y ~ normal(X*beta, sqrt(sigma2));
      target += -log(sigma2);
  }
"

# stan code for global shrinkage with beta-prime prior
code_brg_BetaPrime <- "
  data { 
    int<lower=0> p;
    int<lower=0> n;
    matrix[n,p] X;
    real y[n];
    real<lower=0> a;
    real<lower=0> b;
  }
  parameters {
    vector[p] beta;
    real<lower=0> sigma2;
    real<lower=0> gam2;
    real<lower=0> tau2;
  }
  model {
      gam2 ~ inv_gamma(a, 1);
      tau2 ~ inv_gamma(b, 1/gam2);
      for(i in 1:p){
        beta[i] ~ normal(0, sqrt(sigma2*tau2));
      }
      y ~ normal(X*beta, sqrt(sigma2));
      target += -log(sigma2);
  }
"

# stan code for global shrinkage with inverse-Gaussian prior
code_brg_invGaussian <- "
  data { 
    int<lower=0> p;
    int<lower=0> n;
    matrix[n,p] X;
    real y[n];
    real<lower=0> a;
    real<lower=0> b;
  }
  functions {
    // inverse Gaussian log probability density function
    real igauss_lpdf(real x, real phi, real lambda) {
      return -log(2*pi())/2 + log(lambda)/2 - 3*log(x)/2
             - lambda*square(x - phi)/(2*square(phi)*x);
    }
  }
  parameters {
    vector[p] beta;
    real<lower=0> sigma2;
    real<lower=0> tau2;
  }
  model {
      tau2 ~ igauss(a, b);
      for(i in 1:p){
        beta[i] ~ normal(0, sqrt(sigma2*tau2));
      }
      y ~ normal(X*beta, sqrt(sigma2));
      target += -log(sigma2);
  }
"

# stan code for global shrinkage with Gamma prior
code_brg_Gamma <- "
  data { 
    int<lower=0> p;
    int<lower=0> n;
    matrix[n,p] X;
    real y[n];
    real<lower=0> a;
    real<lower=0> b;
  }
  parameters {
    vector[p] beta;
    real<lower=0> sigma2;
    real<lower=0> tau2;
  }
  model {
      tau2 ~ gamma(a, b);
      for(i in 1:p){
        beta[i] ~ normal(0, sqrt(sigma2*tau2));
      }
      y ~ normal(X*beta, sqrt(sigma2));
      target += -log(sigma2);
  }
"
```

Any of the above stan codes can be compiled using the R package
[rstan](https://cran.r-project.org/web/packages/rstan) and
then used to fit a regression model with global shrinkage priors.
For example:
```{r eval=FALSE}
# library
library(rstan)

# compile stan code 
stan_brg_BetaPrime <- stan_model(model_name = "stan_brg_BetaPrime",
                                 model_code = code_brg_BetaPrime)

# list of input data for rstan
dat <- list(y = y, X = X, n = nrow(X), p = ncol(X), a = 0.5, b = 0.5)

# sample from posterior
fit_rstan <- sampling(stan_brg_BetaPrime, data = dat, iter = 55000,
                      warmup = 5000, thin = 10, verbose = FALSE, chains = 1)

# posterior samples for sigma^2
extract(fit_rstan, "sigma2")$sigma2

# posterior samples for tau^2
extract(fit_rstan, "tau2")$tau2

# posterior samples for betas
extract(fit_rstan, "beta")$beta
```

### Logistic model

Here is an example of stan code to fit a logistic model with a beta prime
prior for global shrinkage.

```{r eval=FALSE}
# logistic regression with beta-prime global shrinkage prior
code_bin_brg_BetaPrime <- "
  data { 
    int<lower=0> p;
    int<lower=0> n;
    matrix[n,p] X;
    int<lower=0, upper=1> y[n];
    real<lower=0> a;
    real<lower=0> b;
  }
  parameters {
    vector[p] beta;
    real<lower=0> gam2;
    real<lower=0> tau2;
  }
  model {
      gam2 ~ inv_gamma(a, 1);
      tau2 ~ inv_gamma(b, 1/gam2);
      for(i in 1:p){
        beta[i] ~ normal(0, sqrt(tau2));
      }
      y ~ bernoulli_logit(X*beta);
  }
"
```
To fit the logistic model with global shrinkage priors the following can be used: 
```{r eval=FALSE}
# library
library(rstan)

# compile stan code 
stan_bin_brg_BetaPrime  <- stan_model(model_name = "stan_bin_brg_BetaPrime",
                                     model_code = code_bin_brg_BetaPrime))

# list of input data for rstan
dat <- list(y = y, X = X, n = nrow(X), p = ncol(X), a = 0.5, b = 0.5)

# sample from posterior
fit_rstan <- sampling(stan_bin_brg_BetaPrime , data = dat, iter = 55000,
                      warmup = 5000, thin = 10, verbose = FALSE, chains = 1)

# posterior samples for tau^2
extract(fit_rstan, "tau2")$tau2

# posterior samples for betas
extract(fit_rstan, "beta")$beta
```

<!----------------------------------------------------------------------------->
<!--------------------------------- LOCAL SHRINKAGE --------------------------->
<!----------------------------------------------------------------------------->
## Regression models with local shrinkage priors

### Gaussian models

```{r eval=FALSE}
# stan code local shrinkage with beta-prime prior
code_brl_BetaPrime <- "
  data { 
    int<lower=0> p;
    int<lower=0> n;
    matrix[n,p] X;
    real y[n];
    real<lower=0> a;
    real<lower=0> b;
    int g[p];
    int<lower=0> K;
  }
  parameters {
    vector[p] beta;
    real<lower=0> sigma2;
    real<lower=0>  gam2[K];
    real<lower=0>  tau2[K];
  }
  model {
      for (k in 1:K) {
        gam2[k] ~ inv_gamma(a, 1);
        tau2[k] ~ inv_gamma(b, 1/gam2[k]);
      }
      for (i in 1:p) {
        beta[i] ~ normal(0, sqrt(sigma2*tau2[g[i]]));
      }
      y ~ normal(X*beta, sqrt(sigma2));
      target += -log(sigma2);
  }
"
```
Again the above stan codes can be compiled using the R package \code{rstan}.
For example:
```{r eval=FALSE}
# library
library(rstan)

# compile stan code 
stan_brl_BetaPrime  <- stan_model(model_name = "stan_brl_BetaPrime",
                                     model_code = code_brl_BetaPrime)))

# list of input data for rstan
dat <- list(y = y, X = X, g = 1:ncol(X), n = nrow(X), p = ncol(X),
            K = ncol(X), a = 0.5, b = 0.5)

# sample from posterior
fit_rstan <- sampling(stan_brl_BetaPrime , data = dat, iter = 55000,
                      warmup = 5000, thin = 10, verbose = FALSE, chains = 1)

# posterior samples for sigma^2
extract(fit_rstan, "sigma2")$sigma2

# posterior samples for tau^2
extract(fit_rstan, "tau2")$tau2

# posterior samples for betas
extract(fit_rstan, "beta")$beta
```


### Logistic model

Here is an example of stan code to fit a logistic model with beta prime
local shrinkage priors.

```{r eval=FALSE}
# stan code local shrinkage with beta-prime prior
code_bin_brl_BetaPrime <- "
  data { 
    int<lower=0> p;
    int<lower=0> n;
    matrix[n,p] X;
    int<lower=0, upper=1> y[n];
    real<lower=0> a;
    real<lower=0> b;
    int g[p];
    int<lower=0> K;
  }
  parameters {
    vector[p] beta;
    real<lower=0>  gam2[K];
    real<lower=0>  tau2[K];
  }
  model {
      for (k in 1:K) {
        gam2[k] ~ inv_gamma(a, 1);
        tau2[k] ~ inv_gamma(b, gam2[k]);
      }
      for (i in 1:p) {
        beta[i] ~ normal(0, sqrt(tau2[g[i]]));
      }
      y ~ bernoulli_logit(X*beta);
  }
"
```
To fit a logistic model with local shrinkage priors use: 
```{r eval=FALSE}
# library
library(rstan)

# compile stan code 
stan_bin_brl_BetaPrime  <- stan_model(model_name = "stan_bin_brl_BetaPrime",
                                     model_code = code_bin_brl_BetaPrime))

# list of input data for rstan
dat <- list(y = y, X = X, g = 1:ncol(X), n = nrow(X), p = ncol(X),
            K = ncol(X), a = 0.5, b = 0.5)

# sample from posterior
fit_rstan <- sampling(stan_bin_brl_BetaPrime , data = dat, iter = 55000,
                      warmup = 5000, thin = 10, verbose = FALSE, chains = 1)

# posterior samples for tau^2
extract(fit_rstan, "tau2")$tau2

# posterior samples for betas
extract(fit_rstan, "beta")$beta
```
